The imprint of global magma oceans on exoplanet demographics

### 2023-02-17 src/tex/ms.tex
draft discussion for statistical imprint

Case~B assumes an inflated radius due to a steam atmosphere for all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$.
This value was found to be a limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
For the rocky interiors, their calculations rely on the same mass-radius relations from ~\citep{Zeng2016} that we applied for case~A-planets.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
We note that an additional radius change due to potentially molten interiors~\citep{Bower2019} is not taken into account here.

@@

As illustrated in Fig.~\ref{fig:radiusevolution}~(bottom panel), different processes can lead to the classification of a planet into either of these categories.
For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a case~C planet, diluting the expected demographic signature.
Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to category~A~\citep[e.g.,][]{Hamano2013}.
To account for these effects, we introduce a free parameter $f_\mathrm{rgh}$ representing the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
%The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
This parametrization introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
In the following, we test if and under what conditions this trend is large enough to be detected with high significance.

@@

\section{Discussion}\label{sec:discussion}
\todo[inline]{\textbf{Potential discussion points: } what is the fraction of \kepler\ planets w/ runaway GH.}

\subsection{Statistical imprint of exoplanet climates}
\todo[inline]{the predicted climate tipping point causes a demographic imprint.\\
...detection would provide an empirical confirmation of the inner edge of the HZ.\\
Discuss our ignorance on $f_\mathrm{rgh}$ and its impact on the imprint.}

@@

>>>>

Case~B assumes an inflated radius due to a steam atmosphere for all planets receiving a dayside-averaged instellation exceeding a threshold of $S_\mathrm{thresh} = \SI{280}{\watt\per\meter\squared}$.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Leconte2013}.
To quantify the radius change, we applied the mass-radius relationships derived by \citet{Turbet2020} using a 1D~inverse radiative-convective model~\citep{Turbet2019}.
For the rocky interiors, their calculations rely on the same mass-radius relations from ~\citep{Zeng2016} that we applied for case~A-planets.
For each planet above the instellation threshold, we assigned the predicted radius for the given water mass fraction and planet mass.
We note that an additional radius change due to potentially molten interiors~\citep{Bower2019} is not taken into account here.

@@

As illustrated in Fig.~\ref{fig:radiusevolution}~(bottom panel), different processes can lead to the classification of a planet into either of these categories.
For instance, both a Hydrogen/Helium-dominated ``sub-Neptune'' and a terrestrial planet currently hosting a steam atmosphere will appear as a case~C planet, diluting the expected demographic signature.
Further, a planet in a runaway greenhouse state will eventually lose its steam atmosphere and move to category~A on a timescale of $10^8\SI{}{\year}$ for a planet with one Earth ocean of water~\citep[][]{Watson1981}.
To account for these effects, we introduce a free parameter $f_\mathrm{rgh}$ representing the fraction of planets above the threshold instellation that are currently inflated due to a runaway greenhouse climate.
Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow relation~A, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow relation~C.
%The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere, or prior complete desiccation.
This parametrization introduces a discontinuity of average planet radii and bulk densities as a function of stellar irradiation.
In the following, we test if and under what conditions this trend is large enough to be detected with high significance.

@@

\section{Discussion}\label{sec:discussion}
\todo[inline]{\textbf{Potential discussion points: }\\
How close is Earth to runaway? Could anthropogenic climate change push us over the edge? (-> \citet{Goldblatt2012}; what is the max. expected change in Earth's albedo due to deglaciation?}

\subsection{Statistical imprint of exoplanet climates}
\todo[inline]{what is the fraction of Kepler planets w/ run-
away GH?}
\begin{note}
The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
Using Bioverse to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we showed that the population of greenhouse-inflated planets create a signature in the radius and density distribution that is well in the detectable range.

However, the idealized step-function central to our hypothesis tests is blurred by several factors.
Its amplitude is affected by ``sub-Neptunes'' with significant Hydrogen/Helium envelopes, which can disguise as inflated rocky planets.
They might not contribute to the demographic signal we are looking for, although a similar transition was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022}.
A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, turning them from planets appearing inflated into much smaller \mbox{(super-)Earths}.
Further dilution comes from the population of planets that lack an atmosphere or sufficient volatiles that can create a steam atmosphere.
There is also a time dependence, both because stellar luminosities change with age and because runaway greenhouse planets with limited water reservoirs ultimately desiccate, ending their steam atmosphere phase.
Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, or surface gravity~\citep{Pierrehumbert2022}.

All of the above mechanisms dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
The ``dilution factor" we introduced embodies our ignorance of these contributions.
Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.

Our finding that the expected trend is strong except for extremely low water mass fractions not only gives reason for optimism regarding its detection, it also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Should a detection be possible, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.

\end{note}

@@

    \todo[inline]{transit depth difference is larger in M dwarfs \citep{Turbet2019}}
